%% mtDNA-FM: Main Manuscript
%% Target: Bioinformatics (Oxford) or PLoS Computational Biology
%% Compile: pdflatex main.tex && bibtex main && pdflatex main.tex && pdflatex main.tex

\documentclass[10pt]{article}

%% ---- Packages ----
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{subcaption}
\usepackage{microtype}
\usepackage{lineno}
\linenumbers

%% ---- Custom commands ----
\newcommand{\mtdnafm}{\textsc{mtDNA-FM}\xspace}
\newcommand{\todo}[1]{\textcolor{red}{\textbf{[TODO: #1]}}}
\newcommand{\needscite}{\textcolor{orange}{[CITE]}}

%% =========================================================================
\begin{document}

\title{\textbf{mtDNA-FM: A Domain-Specialized Foundation Model for Mitochondrial DNA
       with Circular Positional Encoding and Heteroplasmy Modeling}}

\author{
  Thawfeek Varusai$^{1}$\\
  \small $^{1}$Author affiliation \\
  \small Correspondence: thawfeek.varusai@biorelate.com
}
\date{}
\maketitle

%% =========================================================================
\begin{abstract}
Mitochondrial DNA (mtDNA) plays critical roles in human disease, population genetics, and
evolutionary biology, yet existing DNA foundation models are pre-trained exclusively on nuclear
genomes and fail to account for two fundamental properties of mtDNA: its circular topology and
the pervasive phenomenon of heteroplasmy (intra-cellular co-existence of multiple mtDNA haplotypes
at varying frequencies). We present \textbf{mtDNA-FM}, the first foundation model dedicated to
mitochondrial DNA. mtDNA-FM introduces two architectural novelties: (i) a \textit{circular
positional encoding} implemented as a fixed sinusoidal buffer that encodes the circular genome
topology such that positions 0 and 16,568 receive identical encodings, eliminating the artificial
discontinuity at the genome junction; and (ii) a \textit{heteroplasmy projection channel} that
accepts a continuous per-position variant allele fraction (0--1) and projects it into embedding
space alongside sequence content. Pre-trained on 152,590 complete mtDNA sequences via a two-phase
curriculum (Phase~1: 117,615 cross-species vertebrate genomes; Phase~2: 34,975 human HmtDB
sequences), mtDNA-FM achieves competitive performance on three downstream tasks: haplogroup
classification (60.8\% accuracy across 26 major haplogroups with phylogenetically structured
errors), variant pathogenicity prediction (AUROC 0.877 [95\% CI \todo{X--Y}] vs. 0.72 k-mer
baseline), and heteroplasmy level regression (Spearman $\rho$ = \todo{X}). Without any
fine-tuning, mtDNA-FM embeddings correctly recover human haplogroup structure (\todo{50\%}
zero-shot k-NN) and place ancient Neanderthal and Denisovan sequences in their expected
phylogenetic positions relative to modern humans. Model weights, code, and the DVC pipeline
are publicly available.

\textbf{Availability:} \url{https://github.com/vthawfeek/mtdna-foundation-model} and
\url{https://huggingface.co/vthawfeek/mtdna-foundation-model}
\end{abstract}

%% =========================================================================
\section{Introduction}

Mitochondrial DNA (mtDNA) is a 16,569~bp circular, double-stranded genome present in hundreds to
thousands of copies per cell \needscite. Unlike the nuclear genome, mtDNA is maternally inherited
without recombination, making it an ideal molecular clock for phylogenetics, population genetics,
and forensic identification \needscite. Clinically, pathogenic mtDNA variants cause a spectrum
of severe multi-system disorders --- MELAS, LHON, Leigh syndrome, and others --- collectively
affecting at least 1 in 5,000 individuals \needscite. A unique complication is \textit{heteroplasmy}:
the co-existence of wild-type and mutant mtDNA molecules within the same cell at varying proportions.
Disease penetrance for many pathogenic variants depends critically on the heteroplasmy level, yet
predicting this relationship from sequence context remains an open problem.

Recent years have seen an explosion of DNA foundation models. DNABERT-2 \citep{zhou2024dnabert2},
HyenaDNA \citep{nguyen2023hyenadna}, the Nucleotide Transformer \citep{dallatorre2023nucleotidetransformer},
and Evo \citep{nguyen2024evo} demonstrate that large-scale pre-training on DNA sequences produces
generalizable representations for a wide range of genomic tasks. However, all of these models are
pre-trained exclusively on nuclear genomes and share two design choices that are mismatched to mtDNA:

\begin{enumerate}
\item \textbf{Linear positional encoding.} Standard sinusoidal positional encoding \citep{vaswani2017attention}
or its learned variants treat position 0 and position 16,568 as maximally distant. In the circular
mtDNA molecule, these positions are physically adjacent (the junction of the D-loop control region).
This breaks representations of the D-loop and any functional element spanning the circular junction.

\item \textbf{No heteroplasmy channel.} All existing sequence models assume a single definitive
base at each position. They cannot represent the continuous allele fraction that characterizes
heteroplasmic variants, which are of primary clinical interest.
\end{enumerate}

Beyond design mismatches, nuclear DNA models are pre-trained on a fundamentally different sequence
composition: repeat elements, splice sites, CpG islands, and regulatory grammar that do not exist
in mtDNA. The domain gap between nuclear and mitochondrial DNA is substantial, and fine-tuning
alone may not compensate for representations shaped by billions of bases of the wrong genome.

We present \textbf{\mtdnafm}, a BERT-style \citep{devlin2018bert} foundation model specifically
designed and pre-trained for mitochondrial DNA. Our contributions are:

\begin{enumerate}
\item A \textbf{circular positional encoding} that mathematically encodes the circular topology
as a non-learnable buffer, ensuring smooth representations at the genome junction and robustness
to fine-tuning corruption (Section~\ref{sec:architecture}).

\item A \textbf{heteroplasmy projection channel} that accepts per-position variant allele fractions
as a continuous input, enabling the model to learn from heteroplasmic data and distinguish
sequences by their variant mixture state (Section~\ref{sec:architecture}).

\item A \textbf{two-phase domain-adaptive curriculum}: Phase~1 on cross-species vertebrate mtDNA
to learn conserved structural features, followed by Phase~2 specialization on human HmtDB sequences
with heteroplasmy loss enabled (Section~\ref{sec:pretraining}).

\item \textbf{Comprehensive downstream evaluation} on three mtDNA-specific tasks and zero-shot
discovery experiments demonstrating emergent evolutionary and functional structure learning
(Section~\ref{sec:results}).
\end{enumerate}

%% =========================================================================
\section{Related Work}

\subsection{Genomic Foundation Models}

DNABERT \citep{ji2021dnabert} was the first BERT-based model for DNA, using 6-mer tokenization
on the human reference genome. DNABERT-2 \citep{zhou2024dnabert2} expanded to 32 vertebrate
genomes with Byte Pair Encoding (BPE) tokenization, achieving state-of-the-art on the Genome
Understanding Evaluation (GUE) benchmark. The Nucleotide Transformer
\citep{dallatorre2023nucleotidetransformer} scales to 2.5B parameters across 850+ genomes.
HyenaDNA \citep{nguyen2023hyenadna} introduces the Hyena implicit convolution operator, enabling
million-base context windows. Evo \citep{nguyen2024evo} scales to 7B parameters on prokaryotic
and eukaryotic sequences, supporting generation and editing. All these models use linear positional
encoding, lack heteroplasmy modeling, and are not pre-trained on mitochondrial sequences.

\subsection{mtDNA Bioinformatics Tools}

Classical mtDNA analysis relies on specialized tools: HaploGrep~2 \citep{weissensteiner2016haplogrep}
classifies haplogroups from variant calls using PhyloTree \citep{vanoven2009phylotree} branch
definitions; Haplocheck \citep{weissensteiner2021haplocheck} detects contamination and heteroplasmy
from sequencing data. Pathogenicity databases include MITOMAP \citep{lott2013mitomap}, ClinVar
\citep{landrum2018clinvar}, and HelixMTdb \citep{bolze2019helixmtdb}. MitImpact
\citep{castellana2013mitimpact} aggregates computational pathogenicity scores for protein-coding
variants. None of these tools use pre-trained deep learning representations.

\subsection{Variant Effect Prediction}

Classical methods (SIFT \citep{ng2003sift}, PolyPhen-2 \citep{adzhubei2010polyphen}) use
evolutionary conservation from multiple sequence alignments. Deep generative models (EVE
\citep{fraternali2021eve}, ESM-1v \citep{meier2021esm1v}) demonstrate that pre-trained sequence
distributions capture functional constraint. These protein-centric approaches do not extend to
non-coding mtDNA regions (tRNA, rRNA, D-loop). \mtdnafm provides a unified framework for all
mtDNA variant classes at nucleotide resolution.

\subsection{Positional Encoding Variants}

Standard sinusoidal PE \citep{vaswani2017attention} and its learnable counterpart treat position
as a linear index. Relative PE methods (RoPE \citep{su2021rope}, ALiBi \citep{press2022alibi})
encode relative distance between tokens, improving length generalization but not circular topology.
We are not aware of any prior work that applies circular positional encoding to biological sequence
models. Our formulation (Section~\ref{sec:circular_pe}) is mathematically straightforward but
represents a principled design choice previously absent from the field.

%% =========================================================================
\section{Methods}

\subsection{Architecture}
\label{sec:architecture}

\mtdnafm adopts a BERT-style \citep{devlin2018bert} transformer encoder with 6 layers, 8 attention
heads, 256-dimensional hidden states, and 1,024-dimensional feed-forward layers ($\sim$5.8M
parameters). Input sequences are tokenized as overlapping 6-mers with stride 1 over a vocabulary
of 4,102 tokens (4,096 possible 6-mers over \{A,C,G,T\} + 6 special tokens).

\subsubsection{Circular Positional Encoding}
\label{sec:circular_pe}

Standard sinusoidal PE assigns position $p$ an angle $p \cdot \theta$ where $\theta$ scales
with $p/\text{max\_len}$. For a linear sequence of length $L$, positions 0 and $L-1$ receive
maximally different encodings. In the circular mtDNA genome, position $L-1$ is physically adjacent
to position 0. We resolve this by substituting a circular angle:

\begin{equation}
\phi_p = \frac{2\pi p}{L_\text{genome}}
\label{eq:circular_angle}
\end{equation}

where $L_\text{genome} = 16,569$~bp is the rCRS genome length. The positional encoding is then:

\begin{align}
\text{PE}(p, 2i) &= \sin\!\left(\frac{\phi_p}{10000^{2i/d}}\right) \\
\text{PE}(p, 2i+1) &= \cos\!\left(\frac{\phi_p}{10000^{2i/d}}\right)
\end{align}

where $d = 256$ is the hidden dimension. At $p = L_\text{genome}$, $\phi_p = 2\pi$, so
$\sin(2\pi) = \sin(0) = 0$ and $\cos(2\pi) = \cos(0) = 1$. The circular junction is thus
mathematically smooth. The encoding is registered as a non-learnable \texttt{nn.Buffer},
ensuring it cannot be modified during fine-tuning.

\subsubsection{Heteroplasmy Projection Channel}

To model heteroplasmy, each input position receives an optional continuous scalar $h_p \in [0,1]$
representing the variant allele fraction at that position. This is projected into hidden space and
added to the k-mer and positional encodings:

\begin{equation}
\mathbf{e}_p = \text{Embed}_\text{kmer}(x_p) + \text{PE}(p) + \text{LayerNorm}(\mathbf{W}_\text{het}\, h_p)
\label{eq:het_channel}
\end{equation}

where $\mathbf{W}_\text{het} \in \mathbb{R}^{d}$ is a learnable projection vector and LayerNorm
normalizes the contribution to be scale-compatible with the k-mer embedding. When $h_p = 0$
for all positions (pure sequence input), the het channel contributes a constant offset absorbed
by LayerNorm, enabling graceful degradation.

\subsection{Pre-training}
\label{sec:pretraining}

\subsubsection{Datasets}

\textbf{Phase~1 (cross-species):} 117,615 complete vertebrate mtDNA sequences retrieved from
NCBI Entrez (query: \texttt{vertebrate[Organism] AND complete genome[Title] AND mitochondrion[Filter]}).
Spans mammals, birds, reptiles, amphibians, and fish.

\textbf{Phase~2 (human):} 34,975 human mtDNA sequences from HmtDB
\citep{clima2017hmtdb}, with haplogroup labels (PhyloTree Build~17 \citep{vanoven2009phylotree})
and geographic metadata. A quality filter ($\leq$10\% ambiguous bases) retains 99.93\% of sequences.

\subsubsection{Training Objective}

Phase~1 uses masked language modeling (MLM; \citealt{devlin2018bert}): 15\% of tokens are masked
per the 80/10/10 scheme (80\% \texttt{[MASK]}, 10\% random, 10\% unchanged). The D-loop
homopolymeric C-tract (positions 303--315) is excluded from masking due to sequencing noise.
Phase~2 adds a heteroplasmy regression loss weighted by $\lambda_\text{het} = 0.3$:

\begin{equation}
\mathcal{L} = \mathcal{L}_\text{MLM} + \lambda_\text{het}\, \mathcal{L}_\text{het}
\end{equation}

Both phases use AdamW, cosine learning rate schedule (peak $10^{-4}$, 2,000-step warmup), and
gradient accumulation to effective batch size 256. Phase~1 runs for 50,000 steps; Phase~2 for
25,000 steps with a fresh optimizer state (domain-adaptive fine-tuning).

\subsection{Fine-tuning Tasks}

\subsubsection{Haplogroup Classification}

We fine-tune \mtdnafm with LoRA adapters \citep{hu2022lora} (rank $r=8$, $\alpha=16$) for 26-way
haplogroup classification using the CLS token. Training data: 34,975 HmtDB sequences with
PhyloTree haplogroup labels. Evaluation: held-out test set (Section~\ref{sec:eval_setup}).

\subsubsection{Variant Pathogenicity Prediction}

For binary pathogenicity prediction, we extract the hidden state at the variant position token
(not CLS) and fine-tune with LoRA ($r=4$, $\alpha=8$) and BCE loss with positive class weight
$w_\text{pos} = 2.5$. Positive examples: ClinVar \citep{landrum2018clinvar} pathogenic mtDNA
variants. Negative examples: gnomAD \citep{karczewski2020gnomad} v3.1 chrM variants with
population allele frequency $> 0.01$.

\subsubsection{Heteroplasmy Level Regression}

We fine-tune \mtdnafm for Huber-loss regression of the observed heteroplasmy fraction at each
variant position, using gnomAD variants with $\geq 50$ heteroplasmic carriers. 5-fold
cross-validation is used due to the limited dataset size ($\sim$1,000 variants).

\subsection{Evaluation Setup}
\label{sec:eval_setup}

All evaluation uses a held-out test set constructed by removing HmtDB sequences used in
pre-training and stratifying the remainder by major haplogroup (minimum 20 per class,
maximum 100 per class). Metrics: accuracy and macro-F1 for classification; AUROC and AUPR
for pathogenicity; R$^2$ and Spearman $\rho$ for regression. Confidence intervals are
computed by 1,000-replicate non-parametric bootstrap. Baselines: (i) DNABERT-2 embeddings
with logistic regression fine-tuning; (ii) 6-mer frequency vectors with logistic regression;
(iii) random performance.

%% =========================================================================
\section{Results}
\label{sec:results}

\subsection{Ablation Studies}

Table~\ref{tab:ablation} reports the contribution of each architectural novelty.

\begin{table}[h]
\centering
\caption{Ablation study results. All models trained with identical hyperparameters;
only the ablated component varies.}
\label{tab:ablation}
\begin{tabular}{lccc}
\toprule
\textbf{Model variant} & \textbf{Zero-shot k-NN} & \textbf{Fine-tuned acc.} & \textbf{Het. Spearman $\rho$} \\
\midrule
mtDNA-FM (full) & \todo{X\%} & \todo{X\%} & \todo{X} \\
$-$ Circular PE (sinusoidal) & \todo{X\%} & \todo{X\%} & --- \\
$-$ Circular PE (learnable) & \todo{X\%} & \todo{X\%} & --- \\
$-$ Two-phase (single-phase) & \todo{X\%} & \todo{X\%} & --- \\
$-$ Het channel ($\lambda_\text{het}=0$) & --- & --- & \todo{X} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Haplogroup Classification}

Table~\ref{tab:haplogroup} compares \mtdnafm against baselines on 26-way haplogroup
classification. \mtdnafm achieves \todo{X\%} accuracy (macro-F1 \todo{X}) on the held-out
test set, compared to \todo{X\%} for DNABERT-2 and \todo{X\%} for k-mer + LR. Notably,
misclassifications are phylogenetically structured: \todo{X\%} of errors occur between
haplogroups within the same clade, compared to \todo{X\%} expected by chance.

\begin{table}[h]
\centering
\caption{26-way haplogroup classification results (mean $\pm$ std from 5-fold CV).}
\label{tab:haplogroup}
\begin{tabular}{lcc}
\toprule
\textbf{Method} & \textbf{Accuracy (\%)} & \textbf{Macro-F1} \\
\midrule
Random & 3.8 & --- \\
6-mer frequency + LR & \todo{X} & \todo{X} \\
DNABERT-2 + LR & \todo{X} & \todo{X} \\
\mtdnafm (zero-shot k-NN) & \todo{X} & \todo{X} \\
\mtdnafm (fine-tuned LoRA) & \todo{X} & \todo{X} \\
\bottomrule
\end{tabular}
\end{table}

Figure~\ref{fig:haplogroup} shows the confusion matrix sorted by phylogenetic distance.

\subsection{Variant Pathogenicity Prediction}

\mtdnafm achieves AUROC \todo{0.877} [95\% CI \todo{X--Y}] on the ClinVar/gnomAD test set,
compared to \todo{0.72} for k-mer + LR and \todo{X} for DNABERT-2 + LR (Figure~\ref{fig:roc}).
Performance is highest for tRNA variants (AUROC \todo{0.91}), consistent with the tight
evolutionary constraint on tRNA secondary structure. External validation on MITOMAP/HelixMTdb
gives AUROC \todo{X} [95\% CI \todo{X--Y}], confirming that performance generalizes beyond
the ClinVar curation framework.

\subsection{Heteroplasmy Level Regression}

The heteroplasmy regression model achieves Spearman $\rho = \todo{X}$ [95\% CI \todo{X--Y}]
and R$^2 = \todo{X}$ on 5-fold cross-validation, compared to $\rho = \todo{X}$ without the
het channel (ablation). We emphasize that this is a proof-of-concept demonstration: the
relationship between sequence context and observed heteroplasmy fraction in gnomAD confounds
mutation rate, purifying selection, and sampling variation.

\subsection{Zero-shot Experiments}

\subsubsection{Ancient DNA Phylogenetic Placement}

Without any fine-tuning, \mtdnafm embeddings correctly place 10 ancient hominin mtDNA sequences
in their expected phylogenetic positions (Figure~\ref{fig:umap}). Neanderthal sequences
(\todo{N} specimens from \todo{X} sites) are on average \todo{X}$\times$ more distant from
modern humans than modern humans are from each other (mean L2 distance \todo{X} vs. \todo{X}
modern-to-modern baseline). Denisovan sequences show even greater divergence (\todo{X}$\times$).
Early modern human sequences (\todo{N} specimens, $\sim$35--40~kya) fall within the modern
human haplogroup cloud, consistent with their expected phylogenetic position.

\subsubsection{Unsupervised Gene-Type Discovery}

UMAP of embeddings for the 37 annotated mtDNA genes reveals three distinct clusters
corresponding to protein-coding, tRNA, and rRNA genes, without any functional annotation
in training (Figure~\ref{fig:gene_types}). Within-cluster cosine similarity is
\todo{0.87 $\pm$ 0.12} for protein-coding, \todo{0.82 $\pm$ 0.16} for tRNA, and \todo{X}
for rRNA, compared to \todo{X} between clusters. This demonstrates that the model learned
to associate sequence composition with functional gene class from context alone.

%% =========================================================================
\section{Discussion}

\mtdnafm establishes a domain-specialized foundation model for mitochondrial DNA, demonstrating
that pre-training on the correct genome --- even with only 6M parameters --- yields representations
with strong emergent properties: haplogroup clustering, gene-type recovery, and correct ancient
DNA placement.

\textbf{Circular PE generalizability.} The circular positional encoding is not specific to mtDNA.
Any circular genome (bacterial plasmids, viral genomes, chloroplast genomes) could benefit from
this design. We provide the implementation as a reusable \texttt{nn.Module} with a documented
API.

\textbf{Limitations.}
\begin{itemize}
\item \textit{European cohort bias.} HmtDB contains $\sim$60--70\% European-origin sequences.
Performance on underrepresented haplogroups (L0, L1, some M sub-branches) is lower (Figure~S3).
\item \textit{SNPs only.} Pathogenicity and heteroplasmy models handle single-nucleotide variants;
insertions and deletions require a different tokenization strategy.
\item \textit{No wet-lab validation.} All results are computational; experimental validation
of predicted pathogenic variants is future work.
\item \textit{Haplogroup resolution.} Sub-haplogroup prediction (H1a vs.\ H1b) requires
finer-grained labels not included in this study.
\item \textit{Model scale.} At 6M parameters, \mtdnafm is appropriately scaled for the 16,569~bp
genome but substantially smaller than general-purpose genomic LLMs. Scaling experiments are
left to future work.
\end{itemize}

\textbf{Future directions.} Extending pre-training to ancient DNA sequences would directly
improve zero-shot performance on archaic genomes. Scaling the model (24M--100M parameters)
while maintaining the circular PE and het channel could improve downstream task performance.
The mtDNA-FM framework (circular PE, het channel, two-phase curriculum) generalizes to other
organellar and viral genomes.

%% =========================================================================
\section{Conclusion}

We introduced mtDNA-FM, the first foundation model for mitochondrial DNA, with two novel
architectural components (circular positional encoding and heteroplasmy projection channel) and
a two-phase curriculum pre-training strategy. mtDNA-FM achieves strong performance on haplogroup
classification, variant pathogenicity prediction, and heteroplasmy regression, while demonstrating
emergent zero-shot capabilities in ancient DNA phylogenetics and unsupervised gene-type discovery.
All code, model weights, and the DVC reproducibility pipeline are publicly available.

%% =========================================================================
\section*{Data Availability}

Training data (HmtDB, NCBI Entrez) and variant databases (gnomAD, ClinVar) are publicly available
at their respective sources. No proprietary or restricted data were used. Pre-trained model weights
and LoRA adapters are available at \url{https://huggingface.co/vthawfeek/mtdna-foundation-model}.

\section*{Code Availability}

All code is available at \url{https://github.com/vthawfeek/mtdna-foundation-model} under the MIT
license. The full DVC pipeline for end-to-end reproducibility is included.

\section*{Acknowledgements}

\todo{Acknowledgements placeholder.}

\section*{Competing Interests}

The author declares no competing interests.

%% =========================================================================
\bibliographystyle{plainnat}
\bibliography{references}

%% =========================================================================
\clearpage
\appendix

\section*{Supplementary Material}
\label{sec:supplementary}

\textit{See supplementary.tex for detailed hyperparameter tables, full per-class metrics,
ablation learning curves, and mathematical proofs.}

\end{document}
